---
title: Statistic estimation basics
author: ~
date: '2019-11-22'
slug: statistic-estimation-basics
categories: []
tags: []
description: "A basic review of statistics estimation"
output:
  blogdown::html_page:
    highlight: tango
---

# Compare Z-distribution with T-distribution of a estimated mean

We have a sample of a population of mean $\mu$, each of the element of the sample is a random variable with the same probability distribution :

$$X_1, X_2, ..., X_3$$

To estimate the mean of the population from the sample, the sample mean is used : 

$$\bar{X} = \frac{1}{n}\sum_{i=0}^{n}X_i$$

The central limit theorem says that the random variable $\bar{X}$ has a normal distribution of mean :

$$\mu_{\bar{X}} = E\left[\bar{X}\right] = E\left[\frac{1}{n}\sum_{i=0}^{n}X_i\right] = \frac{1}{n} \sum_{i=0}^{n}E\left[X_i\right] = \frac{1}{n}\sum_{i=0}^{n}\mu = \mu $$

and variance :

$$\sigma^{2}_{\bar{X}} = E\left[(\bar{X}-\mu)^2\right] = E\left[(\frac{1}{n}\sum_{i=0}^{n}X_i - \mu)^2\right] = E\left[\frac{1}{n^2}\sum_{i=0}^{n}(X_i - \mu)^2\right] = \frac{1}{n^2}E\left[\sum_{i=0}^{n}(X_i - \mu)^2\right] = \frac{1}{n^2}\sum_{i=0}^{n}E\left[(X_i - \mu)^2\right] = \frac{1}{n^2}\sum_{i=0}^{n}\sigma^2 = \frac{\sigma^2}{n}$$


So the distribution of the random variable :

$$Z = \frac{\bar{X}-\mu}{\frac{\sigma}{\sqrt{n}}}$$

is the standard normal distribution $N(0,1)$.

Unfortunately, the population variance population is rarely available. If the original population is normal or the sample size is large, then the sample variance 

$$S^2 = \frac{1}{n-1}\sum_{i=0}^{n}(X_i - \mu)^2$$

is considered to be a good enough estimator of the population variance. So the random variable 

$$T = \frac{\bar{X_i}-\mu}{\frac{S}{\sqrt{n}}}$$

has a t-distribution with $v=n-1$ degree of freedom. The larger the sample size, the more standard distribution and t-distribution are similar, as shown in the following code and graph.


```{r, warning=FALSE}
library(ggplot2)
library(cowplot)
library(matrixStats)

#set sample size (n), number of sample (n_samples), mean (mu) and sd (sigma)
n = 10
n_samples = 500
mu = 2
sigma = 5

#generate n_samples samples with n values from a normal distribution 
samples <- matrix(rnorm(n*n_samples, mu, sigma), nrow=n_samples ,ncol=n)

#calculate sample mean and sd
x_ = rowMeans(samples)
s = rowSds(samples)

#calculate the values of the Z and T random variables
Z = (x_-mu)/(sigma/sqrt(n))
T = (x_-mu)/(s/sqrt(n))

values = data.frame(cbind(Z,T))

#plot the density of our samples and the true Z-distribution
p1 <- ggplot(values, aes(x = Z)) + 
      stat_density(geom ="line", 
                   position="identity", 
                   color="red", 
                   linetype="dashed") +
      stat_function(fun=dnorm,
                    color="red",
                    args=list(mean=0, 
                              sd=1)) +
      ggtitle("Z-distribution") +
      theme(plot.title = element_text(hjust=0.5, face="bold", size="15"))

#plot the density of our samples and the true T-distribution
p2 <- ggplot(values, aes(x = T)) +
      stat_density(geom ="line", 
                   position="identity", 
                   color="green", 
                   linetype="dashed") + 
      stat_function(fun = dt,
                    color = "green",
                    args = list(df=n-1)) +
      ggtitle("T-distribution") +
      theme(plot.title = element_text(hjust=0.5, face="bold", size="15"))

plot_grid(p1, p2, ncol = 2, nrow = 1)

```

```{r, include=FALSE, warning=FALSE}

library(ggplot2)
library(cowplot)

plot_conf_from_normal <- function(mu, sigma, z_lower_b, z_upper_b, t_lower_b, t_upper_b) {
	cols <- c("Z-statistic"="#f04546","T-statistic"="#3591d1", "Population"="black")
    p <- ggplot(data = data.frame(x=c(mu-3*sigma, mu+3*sigma)), aes(x))  + 
      stat_function(aes(color="Population"),
                    fun=dnorm,
                    args=list(mean=mu, 
                              sd=sigma)) +
      geom_vline(xintercept=mu, color="blue", linetype="dashed") +
      geom_segment(aes(x=z_lower_b,xend=z_upper_b,y=0.002,yend=0.002, color="Z-statistic"), size=1.5)+
      geom_segment(aes(x=z_lower_b,xend=z_lower_b,y=0.0035,yend=0.0005, color="Z-statistic"), size=1.5)+
      geom_segment(aes(x=z_upper_b,xend=z_upper_b,y=0.0035,yend=0.0005, color="Z-statistic"), size=1.5)+
      geom_segment(aes(x=t_lower_b,xend=t_upper_b,y=0.007,yend=0.007, color="T-statistic"), size=1.5)+
      geom_segment(aes(x=t_lower_b,xend=t_lower_b,y=0.0085,yend=0.0055, color="T-statistic"), size=1.5)+
      geom_segment(aes(x=t_upper_b,xend=t_upper_b,y=0.0085,yend=0.0055, color="T-statistic"), size=1.5)+
      scale_colour_manual(name="Legend",values=cols) +
      ggtitle("confidence interval") +
      theme(plot.title = element_text(hjust=0.5, face="bold", size="15"),
            legend.title = element_text(size = 20),
            legend.text = element_text(size = 15))

      return(p)
}

plot_conf_from_binom <- function(p, n, z_lower_b, z_upper_b) {
	cols <- c("Z-statistic"="#f04546", "Population"="black")

	x  <- 0:n
    df <- data.frame(x = x, y = dbinom(x, n, p))

    p <- ggplot(df, aes(x = x, y = y)) + 
         geom_bar(stat = "identity", col = "black", fill = "pink") +
         geom_vline(xintercept=n*p, color="blue", linetype="dashed") +
         geom_segment(aes(x=z_lower_b,xend=z_upper_b,y=0.002,yend=0.002, color="Z-statistic"), size=1.5)+
         geom_segment(aes(x=z_lower_b,xend=z_lower_b,y=0.0035,yend=0.0005, color="Z-statistic"), size=1.5)+
         geom_segment(aes(x=z_upper_b,xend=z_upper_b,y=0.0035,yend=0.0005, color="Z-statistic"), size=1.5)+
         scale_colour_manual(name="Legend",values=cols) +
         ggtitle("confidence interval") +
         theme(plot.title = element_text(hjust=0.5, face="bold", size="15"),
               legend.title = element_text(size = 20),
               legend.text = element_text(size = 15))

      return(p)
}


plot_var_conf_from_normal <- function(mu, sigma, chi_lower_b, chi_upper_b) {
	cols <- c("Chi-statistic"="#f04546", "Population"="black")
    p <- ggplot(data = data.frame(x=c(mu-3*sigma, mu+3*sigma)), aes(x))  + 
      stat_function(aes(color="Population"),
                    fun=dnorm,
                    args=list(mean=mu, 
                              sd=sigma)) +
      geom_vline(xintercept=sigma**2, color="blue", linetype="dashed") +
      geom_segment(aes(x=chi_lower_b,xend=chi_upper_b,y=0.002,yend=0.002, color="Chi-statistic"), size=1.5)+
      geom_segment(aes(x=chi_lower_b,xend=chi_lower_b,y=0.0035,yend=0.0005, color="Chi-statistic"), size=1.5)+
      geom_segment(aes(x=chi_upper_b,xend=chi_upper_b,y=0.0035,yend=0.0005, color="Chi-statistic"), size=1.5)+
      scale_colour_manual(name="Legend",values=cols) +
      ggtitle("confidence interval") +
      theme(plot.title = element_text(hjust=0.5, face="bold", size="15"),
            legend.title = element_text(size = 20),
            legend.text = element_text(size = 15))

      return(p)
}
```

# Estimate a mean with Z and T statistic

```{r}

mu = 10
sigma = 4
n = 10
alpha = 0.99

sample = rnorm(n, mu, sigma)
x_ = mean(sample)
s = sd(sample)

z = qnorm(alpha+(1-alpha)/2)
z_lower_b = x_ - z*(sigma/sqrt(n))
z_upper_b = x_ + z*(sigma/sqrt(n))

t = qt(alpha+(1-alpha)/2, n-1)
t_lower_b = x_ - t*(s/sqrt(n))
t_upper_b = x_ + t*(s/sqrt(n))


cols <- c("Z-statistic"="#f04546","T-statistic"="#3591d1", "Population"="black")

p <- plot_conf_from_normal(mu, sigma, z_upper_b, z_lower_b, t_upper_b, t_lower_b)

p
```

# Estimate the next value

```{r}
mu = 10
sigma = 4
n = 10
alpha = 0.95

sample = rnorm(n, mu, sigma)
x_ = mean(sample)
s = sd(sample)

z = qnorm(alpha+(1-alpha)/2)
z_lower_b = x_ - z*(sigma*sqrt(1+(1/n)))
z_upper_b = x_ + z*(sigma*sqrt(1+(1/n)))

t = qt(alpha+(1-alpha)/2, n-1)
t_lower_b = x_ - t*(s*sqrt(1+(1/n)))
t_upper_b = x_ + t*(s*sqrt(1+(1/n)))


p <- plot_conf_from_normal(mu, sigma, z_upper_b, z_lower_b, t_upper_b, t_lower_b)

p
```

# Estimate the difference between two means

```{r}
mu_1 = 17
mu_2 = 19
sigma_1 = 4
sigma_2 = 4
n_1 = 100
n_2 = 100
alpha = 0.98

sample_1 = rnorm(n_1, mu_1, sigma_1)
sample_2 = rnorm(n_2, mu_2, sigma_2)
x_1 = mean(sample_1)
x_2 = mean(sample_2)
s_1 = sd(sample_1)
s_2 = sd(sample_2)

z = qnorm(alpha+(1-alpha)/2)
z_lower_b = (x_1-x_2) - z*sqrt(sigma_1/n_1+sigma_2/n_2)
z_upper_b = (x_1-x_2) + z*sqrt(sigma_1/n_1+sigma_2/n_2)


v = ((s_1/n_1+s_2/n_2)**2)/(((s_1**2/n_1)**2/(n_1-1))+((s_2**2/n_2)**2/(n_2-1)))
t = qt(alpha+(1-alpha)/2, v)
t_lower_b = (x_1-x_2) - t*sqrt(s_1/n_1+s_2/n_2)
t_upper_b = (x_1-x_2) + t*sqrt(s_1/n_1+s_2/n_2)

p <- plot_conf_from_normal(mu_1-mu_2, sigma_1+sigma_2, z_upper_b, z_lower_b, t_upper_b, t_lower_b)

p
```

# Estimate a proportion p

```{r}

p = 0.4
q = 1- p
n = 30
alpha = 0.99

sample = rbinom(n, 1, p)
p_ = mean(sample)
q_ = (1-p_)


z = qnorm(alpha+(1-alpha)/2)
z_lower_b = p_ - z*sqrt((p_*q_)/n)
z_upper_b = p_ + z*sqrt((p_*q_)/n)

p1 <- plot_conf_from_binom(p, n, n*z_upper_b, n*z_lower_b)

p1
```

# Estimate the difference between two proportions

```{r}
p_1 = 0.4
p_2 = 0.3
q_1 = 1-p_1
q_2 = 1-p_2
n_1 = 30
n_2 = 20
alpha = 0.09

sample_1 = rbinom(n_1, 1, p_1)
sample_2 = rbinom(n_2, 1, p_2)
p_1 = mean(sample_1)
p_2 = mean(sample_2)
q_1 = (1-p_1)
q_2 = (1-p_2)


z = qnorm((1-alpha)+(alpha)/2)
z_lower_b = (p_1-p_2) - z*sqrt(((p_1*q_1)/n_1)+((p_2*q_2)/n_2))
z_upper_b = (p_1-p_2) + z*sqrt(((p_1*q_1)/n_1)+((p_2*q_2)/n_2))

p1 <- plot_conf_from_binom(p_1-p_2, n_1-n_2, (n_1-n_2)*z_upper_b, (n_1-n_2)*z_lower_b)

p1
```

# Estimating the variance

```{r}
mu = 10
sigma = 4
n = 100
alpha = 0.10

sample = rnorm(n, mu, sigma)
x_ = mean(sample)
s = sd(sample)

v = n-1
chi_lower_v = qchisq(alpha/2, v)
chi_upper_v = qchisq(1-(alpha/2), v)

chi_lower_b = (n-1)*(s**2)/(chi_upper_v)
chi_upper_b = (n-1)*(s**2)/(chi_lower_v)



p <- plot_var_conf_from_normal(mu, sigma, chi_upper_b, chi_lower_b)

p2 <- ggplot(data.frame(x = c(0,chi_upper_b+30)), aes(x = x)) +
        stat_function(fun = dchisq, args = list(df = v)) +
        stat_function(fun = dchisq, 
                      args = list(df = v),
                      xlim = c(chi_upper_v,chi_upper_b+30),
                      geom = "area",
                      alpha=0.2) +
        geom_vline(xintercept = chi_upper_v,col="red") +
        stat_function(fun = dchisq, args = list(df = v)) +
        stat_function(fun = dchisq, 
                      args = list(df = v),
                      xlim = c(0,chi_lower_b),
                      geom = "area",
                      alpha=0.2) +
        geom_vline(xintercept = chi_lower_b,col="red") +
        labs(x = "Chi-Squared",y="p")

p; p2

```